Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-04-27 src/docs_bioinference-poster/bioinference-poster.Rmd
Template for #36

* Latent Gaussian models (LGMs) [@rue2009approximate] are three stage hierarchical models with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
* In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs, because ELGM precision matrices are not as sparse
* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
  * Allows a higher degree of non-linearity in the model
* Naomi is an ELGM, not an LGM, because it includes complex dependency structures:
  1. ANC indicators offset from household survey
  2. Incidence depends on prevalence and ART coverage
  3. Observed data are aggregated finer processes
  4. Allow attendance of ART clinics outside home district
* We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools

@@

* **Laplace approximation** Integrate out variables using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
$$
where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
  * Use automatic differentiation via `CppAD` in `TMB`
* **Adaptive Gauss-Hermite Quadrature**
$$
\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
$$
where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ with $k$ points per dimension is adapted based upon
  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$
  * The lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
* Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)
  * Where possible, previously calculated quantities and quadrature rules are reused

```{tikz, fig2, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2.5cm, auto,>=latex', thick, scale = 1]
\node (1) {$p(\theta, x, y)$};
\node (2) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
\node (3) [below of=2] {$\tilde p_\text{LA}(x_i, y)$};
\node (4) [below of=3] {$\tilde p_\text{AQ}(x_i \, | \, y)$};
\node (5) [below of=4] {Output};

\node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
\node (7) [below of=6] {$\tilde p_\text{AQ}(y)$};
\node (8) [below of=7] {$\tilde p_\text{AQ}(\theta \, | \, y)$};
\node (9) [below of=8] {$\tilde p_\text{AQ}(\theta_j \, | \, y)$};
\node (10) [below of=9] {Output};

>>>>

* Latent Gaussian models (LGMs) [@rue2009approximate] are three stage hierarchical models with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
* In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs, because ELGM precision matrices are not as sparse
* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$ and $\mathcal{J_i}$ is some set of indices
  * Allows a higher degree of non-linearity in the model
* Naomi is an ELGM, not an LGM, because it includes complex dependency structures:
  1. ANC indicators offset from household survey
  2. Incidence depends on prevalence and ART coverage
  3. Observed data are aggregated finer processes
  4. Allow attendance of ART clinics outside home district
  5. ART attendance probability as product of prevalence and coverage 
* We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM
* Though we focus on Naomi here, the HIV Inference Group (`hiv-inference.org`) works on many other complex models, challenging for existing Bayesian inference methods, which require flexible modelling tools

@@

* **Laplace approximation** Integrate out variables using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
$$
where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
  * Use automatic differentiation via `CppAD` in `TMB`
* **Adaptive Gauss-Hermite Quadrature**
$$
\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
$$
where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ with $k$ points per dimension is adapted based upon
  * The mode $\hat \theta = \texttt{argmax}_{\theta \in \Theta} p(\theta)$
  * The lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$
* Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)
  * Where possible, previously calculated quantities and quadrature rules are reused

```{tikz, fig2, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2.5cm, auto,>=latex', thick, scale = 1]
\node (1) {$p(\theta, x, y)$};
\node (2) [below of=1] {$\tilde p_\texttt{LA}(x_i, \theta, y)$};
\node (3) [below of=2] {$\tilde p_\texttt{LA}(x_i, y)$};
\node (4) [below of=3] {$\tilde p_\texttt{AQ}(x_i \, | \, y)$};
\node (5) [below of=4] {Output};

\node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
\node (7) [below of=6] {$\tilde p_\texttt{AQ}(y)$};
\node (8) [below of=7] {$\tilde p_\texttt{AQ}(\theta \, | \, y)$};
\node (9) [below of=8] {$\tilde p_\texttt{AQ}(\theta_j \, | \, y)$};
\node (10) [below of=9] {Output};
